Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-01-21 figs/event_types.tex
more figures

\documentclass[10pt,reqno,final]{amsart}
\input{header}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,decorations,calc,math}

\title[Exact Phylodynamics]{Exact Phylodynamics via Structured Markov Genealogy Processes}
\author[King]{Aaron~A.~King}
\address{
  A.~A.~King,
  Department of Ecology \& Evolutionary~Biology,
  Center for the Study of Complex~Systems, and
  Department of Mathematics,
  University of Michigan,
  Ann~Arbor, MI~48109~USA\\
  Santa~Fe~Institute,
  1399 Hyde~Park~Road,
  Santa~Fe, NM~87501~USA
}
\email{kingaa@umich.edu}
\urladdr{\href{https://kinglab.eeb.lsa.umich.edu/}{https://kinglab.eeb.lsa.umich.edu/}}
\author[Lin]{Qianying~Lin}
\address{
  Q.-Y.~Lin,
  Theoretical Biology and Biophysics,
  Los~Alamos National Laboratory,
  Los~Alamos, NM~87545~USA
}
\author[Ionides]{Edward~L.~Ionides}
\address{
  E.~L.~Ionides,
  Department of Statistics
  University of Michigan,
  Ann~Arbor, MI~48109~USA
}
\date{\today}

\hypersetup{pdftitle={Exact Phylodynamics via Structured Markov Genealogy Processes}}
\hypersetup{pdfauthor={A.A. King, Q.-Y. Lin, E.L. Ionides}}
\hypersetup{urlcolor=blue,citecolor=blue,linkcolor=blue,filecolor=blue}

<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
prefix <- "sgp"
source("setup.R")
@
<<packages,include=FALSE,cache=FALSE>>=
library(tidyverse)
library(ggtree)
library(pomp)
library(cowplot)
library(phylopomp)
stopifnot(getRversion() >= "4.3")
stopifnot(packageVersion("pomp")>="5.1")
stopifnot(packageVersion("phylopomp")>="0.9.2")
theme_set(theme_bw(base_family="serif"))
set.seed(1159254136)
@

\begin{document}

\begin{abstract}
  We derive an algorithm for the exact computation of the likelihood of an observed transmission tree.
\end{abstract}

\maketitle

\section{Introduction}

Problem of phylodynamics.
Factorization of problem into two subproblems.

Relation to previous work.
Existing methods \citep{Volz2009a,Stadler2010}.
Large-population, small sample-size approximations.

Extension of previous results \citep{King2022}.
Broader class of state-spaces.
Accommodating discrete structure.

Classes of Markov processes.
Utility and flexibility of Markov assumptions.

Population process induces Markov history and genealogy processes.
Using these, we derive equations for the likelihood of a genealogy conditional on the history.
We then integrate out the history to obtain nonlinear filtering equations, the solution of which yields the likelihood.
These readily lend themselves to a family of sequential Monte Carlo algorithms for computing the likelihood.
We demonstrate with several examples.

In the following, we show a Markov population process of the kind that is a staple in epidemiology induces a Markov process on the space of genealogies.
We then show how one can comput the likelihood of a given genealogy.

\section{Mathematical preliminaries}

\subsection{Notation}

Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\Xr$), denotes a random element.
We will be concerned with a variety of stochastic processes, in both discrete and continuous time.
In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
In the case of continuous-time processes, we will assume that sample paths are right-continuous with left limits (\ie \cadlag).
We will frequently need to refer to the left-limit of such a process.
Accordingly, if $\Zr_t$ is a \cadlag\ random process, we define
\begin{equation*}
  \Zrt_t\colonequals
  \begin{cases}
    \displaystyle\lim_{t'\,\uparrow\,{t}}\;\Zr_{t'}, & t>0,\\[2ex]
    \Zr_0, & t=0.\\
  \end{cases}
\end{equation*}
Note that $\Zrt_t$ is thus left-continuous with right limits.

If $Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its event times in the interval $[0,t]$.

\subsection{Population process}

Motivating examples: compartmental models.
Wide variety of models.
Linear chain trick.
Migration, superspreading, competition between strains.

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \colorlet{darkblue}{blue!50!black}
        \tikzstyle{box}=[draw=black, text=black, fill=white, very thick, minimum size=3em]
        \tikzstyle{ibox}=[draw=darkblue]
        \tikzstyle{deme}=[fill=black!20!white]
        \tikzstyle{label}=[font=\Large]
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{flow}=[draw=black, very thick, >=stealth]
        \tikzstyle{modulate}=[draw=darkblue, >=Circle]

        %% SEIRS model
        \coordinate (origin) at (0,0);
        \node[label] (lab) at (origin) {\textbf{A}};
        \node [box] (S) at ($(origin)+(1,-1)$) {$\lab{S}$};
        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
        \node [box,ibox,deme] (I) at ($(E)+(2,0)$) {$\lab{I}$};
        \node [box] (R) at ($(I)+(2,0)$) {$\lab{R}$};
        \coordinate (overR) at ($(R)+(0,1)$);
        \coordinate (midSE) at ($(S)!0.5!(E)$);
        \draw [flow,->] (S) -- (E);
        \draw [flow,->] (E) -- (I);
        \draw [flow,->] (I) -- (R);
        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
        \draw [modulate,->] (I.north west) .. controls ($(I)+(-1,1)$) and ($(midSE)+(0,1)$) .. (midSE);
        \node[font=\normalsize] (demes) at ($(E)!0.5!(I)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I}\}$};

        %% SIIR (two-strain) model
        \coordinate (origin) at (0,-3);
        \node[label] (lab) at (origin) {\textbf{B}};
        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
        \node [box] (R1) at ($(S)+(6,1)$) {$\lab{R}_1$};
        \node [box] (R2) at ($(S)+(6,-1)$) {$\lab{R}_2$};
        \coordinate (overR) at ($(R1)+(0,1)$);
        \coordinate (underR) at ($(R2)-(0,1)$);
        \node [box,deme] (E1) at ($(S)+(2,1)$) {$\lab{E}_1$};
        \node [box,deme] (E2) at ($(S)+(2,-1)$) {$\lab{E}_2$};
        \node [box,ibox,deme] (I1) at ($(E1)+(2,0)$) {$\lab{I}_1$};
        \node [box,ibox,deme] (I2) at ($(E2)+(2,0)$) {$\lab{I}_2$};
        \draw [flow,->] (S) -- (E1);
        \draw [flow,->] (E1) -- (I1);
        \draw [flow,->] (I1) -- (R1);
        \draw [flow,->] (S) -- (E2);
        \draw [flow,->] (E2) -- (I2);
        \draw [flow,->] (I2) -- (R2);
        \draw [flow,->] (R1) -- (overR) -- (S |- overR) -- (S);
        \draw [flow,->] (R2) -- (underR) -- (S |- underR) -- (S);
        \coordinate (midSE1) at ($(S)!0.5!(E1)$);
        \coordinate (midSE2) at ($(S)!0.5!(E2)$);
        \draw [modulate,->] (I1.north west) .. controls ($(I1)+(-1,1)$) and ($(midSE1)+(0,1.8)$) .. (midSE1);
        \draw [modulate,->] (I2.south west) .. controls ($(I2)+(-1,-1)$) and ($(midSE2)+(0,-1.8)$) .. (midSE2);
        \node[font=\normalsize] (demes) at ($(E2)!0.5!(I2)+(0,-2)$) {$\Demes=\{\lab{E}_1,\lab{I}_1,\lab{E}_2,\lab{I}_2\}$};

        %% COVID model
        \coordinate (origin) at (9,0);
        \node[label] (lab) at (origin) {\textbf{C}};
        \node [box] (S) at ($(origin)+(1,-2)$) {$\lab{S}$};
        \node [box] (R) at ($(S)+(7,1)$) {$\lab{R}$};
        \coordinate (overR) at ($(R)+(0,1)$);
        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
        \node [box,ibox,deme] (Ia) at ($(E)+(2,1)$) {$\lab{I_A}$};
        \node [box,ibox,deme] (Is) at ($(E)+(2,-1)$) {$\lab{I_S}$};
        \node [box] (H) at ($(Is)+(2,0)$) {$\lab{H}$};
        \node [box] (D) at ($(H)+(2,0)$) {$\lab{D}$};
        \draw [flow,->] (S) -- (E);
        \draw [flow,->] (E) -- (Ia.west);
        \draw [flow,->] (E) -- (Is.west);
        \draw [flow,->] (Ia) -- (R);
        \draw [flow,->] (Is) -- (H);
        \draw [flow,->] (H) -- (R);
        \draw [flow,->] (Is) -- (R);
        \draw [flow,->] (H) -- (D);
        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
        \coordinate (midSE) at ($(S)!0.5!(E)$);
        \draw [modulate,->] (Ia.north west) .. controls ($(Ia.north west)+(165:1)$) and ($(midSE)+(0,2)$) .. (midSE);
        \draw [modulate,->] (Is.south west) .. controls ($(Is.south west)+(195:1)$) and ($(midSE)+(0,-2)$) .. (midSE);
        \node[font=\normalsize] (demes) at ($(Is)+(0,-1)$) {$\Demes=\{\lab{E},\lab{I_A},\lab{I_S}\}$};

        %% SI2R (super-spreader) model
        \coordinate (origin) at (9,-5);
        \node[label] (lab) at (origin) {\textbf{D}};
        \node [box] (S) at ($(origin)+(2,-1)$) {$\lab{S}$};
        \node [box,deme] (E) at ($(S)+(2,0)$) {$\lab{E}$};
        \node [box,ibox,deme] (Il) at ($(E)+(2,0)$) {$\lab{I_L}$};
        \node [box,ibox,deme] (Ih) at ($(Il)+(0,-2)$) {$\lab{I_H}$};
        \node [box] (R) at ($(Il)+(2,0)$) {$\lab{R}$};
        \coordinate (overR) at ($(R)+(0,1)$);
        \coordinate (midSE) at ($(S)!0.5!(E)$);
        \draw [flow,->] (S) -- (E);
        \draw [flow,->] (E) -- (Il);
        \draw [flow,<->] (Il) -- (Ih);
        \draw [flow,->] (Il) -- (R);
        \draw [flow,->] (Ih.east) -- (R |- Ih.east) -- (R);
        \draw [flow,->] (R) -- (overR) -- (S |- overR) -- (S);
        \draw [modulate,->] (Il.north west) .. controls ($(Il.north west)+(135:0.8)$) and ($(midSE)+(0,1)$) .. (midSE);
        \draw [modulate,->] (Ih.west) .. controls ($(Ih.west)+(180:0.8)$) and ($(midSE)+(0,-1)$) .. (midSE);
        \node[font=\normalsize] (demes) at (midSE |- Ih) {$\Demes=\{\lab{E},\lab{I_L},\lab{I_H}\}$};
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Examples of discretely-structured population models.
    Demes are shaded.
    \textbf{(A)} An SEIRS model.
    Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
    Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
    If this immunity wanes, they re-enter the susceptible compartment.
    Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
    Accordingly, there are two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
    If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
    \label{fig:examples}
  }
\end{figure}

We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}
  \label{eq:ic}
  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
We assume that $\Kr_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
  \end{gathered}
\end{equation*}
We will further assume that $\Xr_t$ is non-explosive, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t$.

\subsubsection{Kolmogorov forward equation}

The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
\begin{equation}
  \label{eq:kfe}
  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
\end{equation}
and if, moreover, $w(0,x) = p_0(x)$,
then $\int_{\mathcal{E}}\!{w(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.

Another perspective on the Markov processes is to be had from its Markov state transition diagram (\cref{fig:markov_state}).

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \usetikzlibrary{shapes,arrows,positioning}
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum height=8em]
        \tikzstyle{trans}=[color=black, font=\Large, thick, >=stealth]
        \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
        \node[state] (trans) at (165:8) {$x'=(S-1,E+1,I,R)$};
        \node[state] (prog) at (205:7) {$x'=(S,E-1,I+1,R+1)$};
        \node[state] (recov) at (305:5) {$x'=(S,E,I-1,R+1)$};
        \node[state] (wane) at (350:8) {$x'=(S+1,E,I,R-1)$};
        \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
        \draw[trans,->] (base) -- (trans) node[midway,above,sloped] {$\lab{Trans}$};
        \draw[trans,->] (base) -- (prog) node[midway,above,sloped] {$\lab{Prog}$};
        \draw[trans,->] (base) -- (recov) node[midway,above,sloped] {$\lab{Recov}$};
        \draw[trans,->] (base) -- (wane) node[midway,above,sloped] {$\lab{Wane}$};
        \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
        \node[font=\LARGE] (U) at (95:3.5) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Markov state transition diagram for the SEIRS model depicted in \cref{fig:examples}A.
    The state, $x$, is characterized by four numbers, $S$, $E$, $I$, and $R$.
    From a given state $x$, there are five possible kinds of jumps $x\mapsto{x'}$.
    Accordingly, the set, $\Jumps$, of jump marks has five elements.
    Each of these is of a different type:
    $\lab{Trans}$ (transmission) is of birth type,
    $\lab{Prog}$ (progression) is of migration type,
    $\lab{Recov}$ (recovery) is of death type,
    $\lab{Sample}$ (sampling) is of sample type,
    and $\lab{Wane}$ (loss or waning of immunity) is of neutral type.
    See \cref{sec:jump_types} for a description of these jump types.
    Note that, in this formulation, when a sampling event occurs, the state does not change.
    \label{fig:markov_state}
  }
\end{figure}

\subsubsection{Structured populations, demes, and deme occupancy}
\label{sec:demes}

In an \emph{unstructured} Markov population process, every lineage is exactly like every other.
\citet{King2022} showed how every such process induces an unstructured Markov genealogy process.
Here, our aim is to expand the theory considerably by allowing our population of lineages to have discrete structure.
In particular, we suppose that there are a countable set of subpopulations that differ in their vital rates, but within each of which, individual lineages are statistically identical.
We call these subpopulations \emph{demes}, and use the symbol $\Demes$ to denote an index set for them.

For any $i\in\Demes$, we let $n_i(\Xr_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
Thus $n(\Xr_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.

\subsubsection{Jump marks}

In the following, it will be useful to break the jumps into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
In \cref{fig:markov_state}, we use the marks to distinguish events with distinct effects on genealogical relationships.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

Let us define the \emph{jump mark} process, $\Ur_t$, to be the mark of the latest jump as of time $t$.
As usual, we take the sample paths of $\Ur_t$ to be \cadlag.
Observe that $(\Xr_t,\Ur_t)$ is a Markov process, though $\Ur_t$ is not.

\subsection{Examples}

\subsubsection{SEIRS model}

\subsubsection{SIIR model}

\subsubsection{Linear birth-death model}

\subsubsection{Moran model and the Kingman coalescent}

\subsection{History process}

Consider the Markov process $(\Xr_t,\Ur_t)$.
We define its \emph{history process}, $\Hr_t$, to be the restriction of the random function $s\mapsto(\Xr_s,\Ur_s)$ to the interval $[0,t]$.
Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
Alternatively, one can think of $\Hr_t$ as consisting of the sequence
$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
\begin{equation*}
  \label{eq:Hdens}
  \begin{aligned}
    \pi^{\mathrm{H}}(\dd{\H_t})
    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
  \end{aligned}
\end{equation*}
where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.

\subsection{Binomial ratio}
\label{sec:binomial_ratio}

For $n,r,\ell,s\in{\Zp^\Demes}$, define the \emph{binomial ratio}
\begin{equation*}
  \BinRatio{n}{\ell}{r}{s}\colonequals
  \begin{cases}
    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
         & \text{if}\ \forall i\ n_i\ge{\Set{\ell_i,r_i}}\ge{s_i}\ge{0},\\
         0, & \text{otherwise}.
  \end{cases}
\end{equation*}
Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
Moreover, in consequence of the Chu-Vandermonde identity, we have
\begin{equation*}
  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
\end{equation*}
whenever $n_i\ge{\Set{\ell_i,r_i}}\ge{0}$ for all $i$.

\section{The induced genealogy process}

\subsection{Genealogies}

For our purposes, a \emph{genealogy}, $G_{T}$, is a labeled, time-calibrated tree.
It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
There are three distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled living lineages;
\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
\item \emph{sample nodes}, which represent labeled samples.
\end{inparaenum}
Each node $a$ is associated with a specific time, $t(a)$.
In particular, if $a$ is a tip node in $G_T$, then $t(a)=T$;
if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
We let $\event{G}{T}$ denote the set of all node-times of the genealogy $G_T$.
Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.

Formally, we define a genealogy, $G_T$, to be an ordered pair $(y,z)$.
The $z$ element specifies the topological structure, as follows.
Let $\leaves{G_T}$ be the set of all tip and sample nodes in $G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{G_T}$.
The partition \emph{fineness} defines a partial order on $\func{P}$.
Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
The topological structure of $G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.

The $y$ element of $G_T$ specifies the coloring of branches and location of nodes.
In particular, for $a\in\leaves{G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
Thus $d:[0,T]\times\leaves{G_T}\to\Demesplus$ is a well-defined \cadlag\ function.

Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
Let $m:[0,T]\times\leaves{G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
Clearly, $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.

Finally, we define the map $y:[0,T]\times\leaves{G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
The two \cadlag\ functions $y$ and $z$ specify the genealogy $G_T$ completely and uniquely.
When the genealogy is not clear from context, we write $y^{G}$ and $z^{G}$ to refer to the respective components of genealogy $G$.

Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
In particular, $y$ must respect the topological structure of $G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
In other words, if $a,b\in{c}\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
Let $\Yspace^z$ denote the space of all functions that respect the topological structure described by $z$.
Let $\Yspace^\z_t$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.

\subsection{Jump types}
\label{sec:jump_types}

\begin{figure}
  \begin{center}
    \resizebox{0.8\linewidth}{!}{
      \begin{tikzpicture}[scale=2]
        \usetikzlibrary{shapes,arrows,positioning}
        \usetikzlibrary{arrows.meta,positioning,decorations}
        \colorlet{darkblue}{blue!50!black}
        \colorlet{maize}{yellow!90!black}
        \colorlet{grey}{black!30!white}
        \colorlet{darkgreen}{green!40!black}
        \colorlet{lightblue}{white!30!blue}
        \tikzstyle{label}=[font=\Large]
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{deme1}=[color=grey, line width=3pt, >={Circle[width=5pt,length=5pt]}]
        \tikzstyle{deme2}=[color=darkblue, line width=3pt, >={Circle[width=5pt,length=5pt]}]
        \tikzstyle{deme3}=[color=maize, line width=3pt, >={Circle[width=5pt,length=5pt]}]
        \tikzstyle{axis}=[color=black, thick, >=stealth]
        \tikzstyle{timepoint}=[color=red!70!black]

@@

        %% Birth-type event
        \begin{scope}[shift={(0,0)}]
          \coordinate (origin) at (0,0);
          \node[label] (lab) at (origin) {\textbf{A}};
          \coordinate (left) at ($(origin)+(0.1,-0.1)$);
          \coordinate (right) at ($(left)+(1,0)$);
          \coordinate (mid) at ($(left)!0.5!(right)$);
          \coordinate (rlab) at ($(right)+(0.1,0)$);
          \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
          \coordinate (bright) at (rlab |- bleft);
          \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
          \coordinate (temp1) at ($(temp)+(0,\ysp)$);
          \coordinate (temp2) at ($(temp)-(0,\ysp)$);
          \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
          \draw[deme3] (mid |- temp) -- (mid |- temp2) -- (right |- temp2);
          \draw[deme1] (left |- temp) -- (right |- temp);
          \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
          \foreach \y in {1,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme1] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {6,8} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme2] (left |- temp) -- (right |- temp);
          }
          \foreach \y in {2,7} {
            \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
            \draw[deme3] (left |- temp) -- (right |- temp);
          }
          \draw[timepoint] (mid) -- (mid |- bleft);
          \draw[axis,->] (bleft) -- (bright);
          \node[xshift=1ex] (tlab) at (bright.east) {$t$};
        \end{scope}

>>>>

\begin{center}
  \resizebox{0.8\linewidth}{!}{
    \begin{tikzpicture}[scale=2]
      \usetikzlibrary{shapes,arrows,positioning}
      \usetikzlibrary{arrows.meta,positioning,decorations}
      \tikzstyle{label}=[font=\Large]
      \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
      \tikzstyle{deme1}=[color=darkblue, line width=3pt]
      \tikzstyle{deme1H}=[color=darkblue, line width=3pt, densely dotted]
      \tikzstyle{deme2}=[color=maize, line width=3pt]
      \tikzstyle{deme2H}=[color=maize, line width=3pt, densely dotted]
      \tikzstyle{axis}=[color=black, thick, >=stealth]
      \tikzstyle{timepoint}=[color=red!70!black]

@@

      %% Birth-type event
      \begin{scope}[shift={(0,0)}]
        \coordinate (origin) at (0,0);
        \node[label] (lab) at (origin) {\textbf{A}};
        \coordinate (left) at ($(origin)+(0.1,-0.1)$);
        \coordinate (right) at ($(left)+(1,0)$);
        \coordinate (mid) at ($(left)!0.5!(right)$);
        \coordinate (rlab) at ($(right)+(0.1,0)$);
        \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
        \coordinate (bright) at (rlab |- bleft);
        \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
        \coordinate (temp1) at ($(temp)+(0,\ysp)$);
        \coordinate (temp2) at ($(temp)-(0,\ysp)$);
        \draw[deme2] (mid |- temp) -- (mid |- temp1) -- (right |- temp1);
        \draw[deme1] (mid |- temp) -- (mid |- temp2) -- (right |- temp2);
        \draw[deme1] (left |- temp) -- (right |- temp);
        \node[text=darkgreen,font=\Huge] at (mid |- temp) {$\bullet$};
        \foreach \y in {1,7} {
          \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
          \draw[deme1] (left |- temp) -- (right |- temp);
        }
        \foreach \y in {6,8} {
          \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
          \draw[deme2] (left |- temp) -- (right |- temp);
        }
        \foreach \y in {2,7} {
          \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
          \draw[deme2] (left |- temp) -- (right |- temp);
        }
        \draw[timepoint] (mid) -- (mid |- bleft);
        \draw[axis,->] (bleft) -- (bright);
        \node[xshift=1ex] (tlab) at (bright.east) {$t$};
      \end{scope}

      %% Death-type event
      \begin{scope}[shift={(2,0)}]
        \coordinate (origin) at (0,0);
        \node[label] (lab) at (origin) {\textbf{B}};
        \coordinate (left) at ($(origin)+(0.1,-0.1)$);
        \coordinate (right) at ($(left)+(1,0)$);
        \coordinate (mid) at ($(left)!0.5!(right)$);
        \coordinate (rlab) at ($(right)+(0.1,0)$);
        \coordinate (bleft) at ($(left)+(0,-8*\ysp)$);
        \coordinate (bright) at (rlab |- bleft);
        \foreach \y in {1,2,6} {
          \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
          \draw[deme1] (left |- temp) -- (right |- temp);
        }
        \foreach \y in {3,8} {
          \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
          \draw[deme2] (left |- temp) -- (right |- temp);
        }
        \coordinate (temp) at ($(origin)+(0,-4*\ysp)$);
        \draw[deme2H] (left |- temp) -- (mid |- temp);
        \node[text=red,font=\Huge] at (mid |- temp) {$\times$};
        \foreach \y in {5,7} {
          \coordinate (temp) at ($(origin)+(0,-\y*\ysp)$);
          \draw[deme2] (left |- temp) -- (right |- temp);
        }
        \draw[timepoint] (mid) -- (mid |- bleft);
        \draw[axis,->] (bleft) -- (bright);
        \node[xshift=1ex] (tlab) at (bright.east) {$t$};
      \end{scope}
